---
title: "Class 15: Transcriptonomics and analysis of RNA-Seq data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import countData and colData into R

```{r}
counts <- read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)
```

look at data

```{r}
dim(counts)
head(counts) #shows first six rows
```

```{r}
dim(metadata)
head(metadata)
```

make sure column names count data is the first column of metadata
```{r}
colnames(counts)
```

```{r}
metadata$id
```

check for corespondence of metadata and countdata
```{r}
 #everything but first one 
#all() function makes sure ALL are true. if one is false, will return FALSE
all( colnames(counts)[-1] == metadata$id )
```
 **side note: View() function in Console allows data to be viewed in new tab**
 
First task- find which columns in **counts** correspond to the control conditions (no drug)
Then calculate the mean value for each gene.

```{r}
control.inds <- metadata$dex == "control"
control.inds <- metadata$id[control.inds]

head( counts[,control.inds] )
```

Calculate the mean counts value for all genes
```{r}
#rowSums(counts[,control.inds]) / 4 ...not a good way in case you get more columns (perform more experiments)...four is hard coded into the code

control.mean <- rowSums(counts[,control.inds]) / ncol(counts[,control.inds])
names(control.mean) <- counts$ensgene
head(control.mean)
```

Now do the same thing for the drug treated samples (*treated* columns)

```{r}

treated.inds <- metadata$dex == "treated"
treated.inds <- metadata$id[treated.inds]

head( counts[,treated.inds] )
```

```{r}
treated.mean <- rowSums(counts[,treated.inds]) / ncol(counts[,treated.inds])
names(treated.mean) <- counts$ensgene
head(treated.mean)
```

Combine the meancount data for control and treated into a new data.frame called meancounts
```{r}
meancounts <- data.frame(control.mean, treated.mean)

```

```{r}
colSums(meancounts)
```

```{r}
plot(meancounts$control.mean, meancounts$treated.mean)
```

Put this on a log log plot to help see the data!
```{r}
plot(meancounts$control.mean, meancounts$treated.mean, log = "xy")
```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

note: the NaN (“not a number””) and -Inf (negative infinity) results
NaN is returned when you divide by zero and try to take the log 
-Inf is returned when you try to take log of zero. It turns out that there are a lot of genes with zero expression...filter data to remove these genes.


```{r}
#practice vector
toy <- c(1,2,0,4,0)
toy == 0
which(toy == 0)
```

```{r}
toy <- meancounts[1:6,]

unique( which(toy[,1:2] == 0, arr.ind = TRUE)[,1] )
```

Remove the zero entry genes from meancounts to produce **mycounts**
```{r}
zero.vals <- which(meancounts[,1:2] == 0, arr.ind = TRUE)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)
```
```{r}
sum(up.ind)
sum(down.ind)
```
note: back ticks (``) run code as R code when document is knitted!

There are `sum(up.ind)` up regulated genes and `sum(down.ind)` down regulated genes.

```{r}
head(mycounts[up.ind,])
```

need to put data back into "human speak"...biology terms

##Adding Annotation Data

Use the **merge()** function to add annotation data from ensemble

```{r}
anno <- read.csv("data/annotables_grch38.csv")
head(anno)
```

```{r}
mycounts.annotated <- merge(x = mycounts, y = anno, by.x = "row.names", by.y = "ensgene")
head(mycounts.annotated)
#now we know what the genes are!
```

## DESeq2 analysis

Now we do this analysis properly with p-values!

```{r}
library(DESeq2)
citation("DESeq2")
```
```{r}
dds <- DESeqDataSetFromMatrix(countData = counts, 
                              colData = metadata, 
                              design = ~dex, 
                              tidy = TRUE)
dds
```

```{r}
dds <- DESeq(dds)
```

get the results
```{r}
res <- results(dds)
head(res)
```

```{r}
summary(res)
```

'''
```{r}
res$sig <- res$padj<0.05
```

```{r}
#mycols <- rep("gray", nrow(res01))
#mycols[ abs(res01$log2FoldChange) > 2 ] <- "red"

#inds <- (res01$padj < 0.01) & (asbs(res01$log2FoldChange) > 2)
#mycols[inds] <- "blue"
#plot(res01$log2FoldChange, -log(res01$padj), col = mycols, ylab = "-Log(P-value", xlab = "Log2(FoldChange)")
#abline(v = c(-2,2), col = "gray", lty = 2)
#abline(h = -log(0.1), col = "gray", lty = 2)
```

```{r}
library(ggplot2)

ggplot(as.data.frame(res), aes(log2FoldChange, -1*log10(pvalue), col = sig)) + 
    geom_point() + 
    ggtitle("Volcano plot")
```

















