---
title: "Class 11: Structural Bioinformatics 1"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Section 1- The PDB

```{r}
stats <- read.csv("class11 pdb data csv.csv", row.names = 1)
stats
```

Q1:  Determine the percentage of structures solved by X-Ray and Electron Microscopy. From
the website what proportion of structures are protein?

```{r}
percent.by.method <- stats$Total / sum(stats$Total) * 100
names(percent.by.method) <- rownames(stats)
percent.by.method
```

```{r}
sum(stats$Proteins) / sum(stats$Total) * 100
```

Type HIV in the PDB website search box on the home page and determine how many
HIV-1 protease structures are in the current PDB?
  -> 1138
  
  
## load the Bio3D package (previously installed...don't need to install again)

```{r}
library(bio3d)  
```

More about the functions within the package:
```{r}
help(package = bio3d)
example(plot.bio3d)
```

Reading pdb files:

```{r}
pdb <- read.pdb("1hsg")
## Note: Accessing on-line PDB file
pdb
```
```{r}
pdb$atom
```
```{r}
pdb$atom$resid
```

```{r}
aa321(pdb$atom$resid)
```

Goint to use:
  pdb <- read.pdb("1hsg")
  prot.inds <- atom.select(pdb, ?)
  lig.inds <-

```{r}
?atom.select
```

Want to select out the protein and drug only parts of these molecular files.

```{r}
pdb <- read.pdb("1hsg")
prot.inds <- atom.select(pdb, "protein")
```

```{r}
pdb$atom[prot.inds$atom,]
```

```{r}
prot.pdb <- trim.pdb(pdb, prot.inds)
write.pdb(prot.pdb, file = "protein.pdb")
```

Let's do the same thing for our drug...

```{r}
lig.inds <- atom.select(pdb, "ligand")
lig.pdb <- trim.pdb(pdb, lig.inds)
write.pdb(lig.pdb, file = "ligand.pdb")
```

```{r}
library(bio3d.view)
#view(pdb, "overview")
```

## Bio3D view() installed...for the 3D viewer in your R markdown (the development version of bio3d.view)

in r console:
> install.packages("devtools")
> devtools::install_bitbucket("Grantlab/bio3d-view")

To use in your R session:

```{r}
library("bio3d.view")
pdb <- read.pdb("5p21")
#view(pdb)
#view(pdb, "overview", col = "sse")
```




