---
title: "Class 13: Structure Based Drug Discovery"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Prepare HIV-protease molecule for docking

First get a PDB structure from the online database then clean it up for docking

```{r}
library(bio3d)
```

get protein structure from the database
```{r}
file.name <- get.pdb("1hsg")
```

```{r}
file.name
```

Read file into R and clean it up (remove non-protein bits) and save ligand into a separate file

```{r}
hiv <- read.pdb(file.name)
hiv
```

Q1: What is the name of the two non protein resid values in this structure? What does resid
correspond to and how would you get a listing of all reside values in this structure? 
  -> HOH & MK1
  
```{r}
prot <- trim.pdb(hiv, "protein") #no water
prot
```

```{r}
write.pdb(prot, file = "1hsg_protein.pdb")
```


```{r}
lig <- trim.pdb(hiv, "ligand") #ligand is MK1
lig
```

```{r}
write.pdb(lig, file = "1hsg_ligand.pdb")
```

##Inspecting docking results
```{r}
res <- read.pdb("all.pdbqt", multi = TRUE)
write.pdb(res, "results.pdb")
res
```

## Normal Mode Analysis

```{r}
pdb <- read.pdb("1HEL")

m <- nma(pdb)
plot(m)
```

```{r}
mktrj(m, mode = 7, file = "nma_7.pdb")
```

Can now run this on VMD, and play to show predicted flexibility of the molecule!


