---
title: "Class 6"
author: "Madison Hale"
date: "1/24/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Section 1: Reading Files again :)

Going to use the **read.table()** function and friends to read example flat files

First, we add a simple plot:

```{r}
plot(1:20)
```

Back to file reading...

```{r}

read.table("https://bioboot.github.io/bimm143_W19/class-material/test1.txt", header = TRUE, sep = ",")

```

For this common CSV format we can use ""read.csv()**

```{r}

data1 <- "https://bioboot.github.io/bimm143_W19/class-material/test1.txt"
data1 <- read.csv(data1)
data1

```

```{r}
data2 <- "https://bioboot.github.io/bimm143_W19/class-material/test2.txt"
data2 <- read.csv(data2, sep = "$")
data2

```

```{r}

data3 <- "https://bioboot.github.io/bimm143_W19/class-material/test3.txt"
data3 <- read.table(data3, sep = "")
data3

```

### Section 2: R Functions

My first function:

```{r}

add <- function(x, y=1) {
 # Sum the input x and y
 x + y
}

```

Using this function

```{r}

add(1)

```

```{r}

add(1,100)

```

```{r}

add( c(1, 2, 3) )

```

```{r}

add( c(1, 2, 3), 4 )

```

some improper add() use: 

```{r}

#add(1, 2, 2) 

```

```{r}

#add(x=1, y=“b”)

```

When would you write a function? ...when you find yourself doing the same thing 3+ times

My second function: 

```{r}

rescale <- function(x) {
 rng <- range(x)
 (x - rng[1]) / (rng[2] - rng[1])
}

```

testing this function on a small sample, where we know what answer SHOULD be:

```{r}

rescale(1:10)

```

Test, **Fail**, Change, Test again,…

```{r}

# How would you get your function to work here…
rescale( c(1,2,NA,3,10) )

```

```{r}

# What should your function do here?
#rescale( c(1,10,"string") )

```

```{r}

rescale2 <- function(x) {
 rng <- range(x, na.rm = TRUE)
 (x - rng[1]) / (rng[2] - rng[1])
}

x <- (c(1,2,NA,3,10) )
rng <- range(x)
#rng

```

```{r}

rescale2( c(1,2,NA,3,10))

```

Trying another function...

```{r}

rescale3 <- function(x, na.rm=TRUE, plot=FALSE) {
 rng <-range(x, na.rm=na.rm)
 print("Hello")
 answer <- (x - rng[1]) / (rng[2] - rng[1])
 print("is it me you are looking for?")
 if(plot) {
 plot(answer, typ="b", lwd=4)
   print("please don't ever sing again ;)")
 }
 print("I can see it in ...")
 return(answer)
}

```


using it

```{r}

rescale3( c(1:6, NA, 10), plot = TRUE)

```

### Working with the bio3d package

To install this package I used the command **install.packages("bio3d")**

Time to use it! :)

need to call library function to use the functions within the package...
```{r}

library(bio3d)

```

```{r}
# Read a PDB file from the database

s1 <- read.pdb("4AKE") # kinase with drug
s1

```

Let's improve this code!

```{r}

# Can you improve this analysis code?
library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain = "A", elety = "CA")
s2.chainA <- trim.pdb(s2, chain = "A", elety = "CA")
s3.chainA <- trim.pdb(s3, chain = "A", elety = "CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse = s1.chainA, typ = "l", ylab = "Bfactor")
plotb3(s2.b, sse = s2.chainA, typ = "l", ylab = "Bfactor")
plotb3(s3.b, sse = s3.chainA, typ = "l", ylab = "Bfactor")


```

HOMEWORK:
find the smallest working snipet of code, make it as simple as possible
wrap it into a function
...see if it works!
Then, you'd be able to run it on any protein structure!


Cleaning up the above code:

```{r}

library(bio3d)
s1 <- read.pdb("4AKE") # kinase with drug
s2 <- read.pdb("1AKE") # kinase no drug
s3 <- read.pdb("1E4Y") # kinase with drug

s1.chainA <- trim.pdb(s1, chain = "A", elety = "CA")
s2.chainA <- trim.pdb(s2, chain = "A", elety = "CA")
s3.chainA <- trim.pdb(s3, chain = "A", elety = "CA")

s1.b <- s1.chainA$atom$b
s2.b <- s2.chainA$atom$b
s3.b <- s3.chainA$atom$b

plotb3(s1.b, sse = s1.chainA, typ = "l", ylab = "Bfactor")
plotb3(s2.b, sse = s2.chainA, typ = "l", ylab = "Bfactor")
plotb3(s3.b, sse = s3.chainA, typ = "l", ylab = "Bfactor")

#Simplify to work with some generic vectors
#s <- read.pdb("input kinase")
#chain <- trim.pdb(s, chain ="A", elety ="CA")
#s.b <- c(s1.chainA$atom$b, s2.chainA$atom$b, s3.chainA$atom$b)
#plots <- c(plotb3(s1.b, sse = s1.chainA, typ = "l", ylab = "Bfactor"), plotb3(s2.b, sse = s2.chainA, typ = "l", ylab = "Bfactor"), plotb3(s3.b, sse = s3.chainA, typ = "l", ylab = "Bfactor"))


  
rescale <- function(x) {
 rng <- range(x)
 (x - rng[1]) / (rng[2] - rng[1])
}




```





